---
title: "Get started with birdnetR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with birdnetR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(birdnetR)
```

The `birdnetR` package provides a comprehensive interface for utilizing the `birdnet` Python package within R. This guide will walk you through the basic steps of setting up the package, initializing models, and using various functions to analyze audio files for (bird) species identification.


##  Installation


First, install the `reticulate` package if you haven't already:

```{r install_reticulate}
install.packages("reticulate")
```

For now, you have to install the package from GitHub. 

To install the package directly from GitHub, you need to have the devtools package installed. If you don't have devtools installed, you can install it using:

```{r install_devtools}
install.packages("devtools")
```

Then, you can install `birdnetR` from GitHub:

```{r install_birdnetR}
devtools::install_github("birdnet-team/birdnetR")
```

Next, install `birdnet`, which will set up a Python virtual environment named `r-birdnet` by default. You can configure this with the envname parameter. Do this only once during the initial setup or if you encounter issues with the environment.
```{r load_and_install_birdnet}
library(birdnetR)
install_birdnet()
```


<div style="padding: 15px; margin-bottom: 20px; border: 1px solid #bce8f1; border-radius: 4px; background-color: #d9edf7; color: #31708f;">
<strong>Incompatible Python Version</strong><br>
If you do not have a compatible Python version installed, you can use <code>reticulate::use_python_version(version = &lt;version&gt;)</code>.
</div>

### Custom virtual environment
By default, a virtual environment named `r-birdnet` will be used. If you don't need any customization, you can ignore this section, and everything should work out-of-the-box.

If you need to use a custom name for the virtual environment, you can specify it using the `envname` argument:

```{r}
library(birdnetR)
install_birdnet(envname = "custom")
```

The next time you start a new R session, tell `reticulate` to use this environment after loading your libraries:
```{r}
library(birdnetR)
reticulate::use_virtualenv("custom")
```

By default, `reticulate` will install to and use a folder `.virtualenvs` in your home directory. You can also specify paths to your virtual environments, which is useful if you want to install `birdnetR` into an existing virtual environment:

```{r}
library(birdnetR)
path_venv <- "/path/to/existing/venv"
install_birdnet(envname = path_venv)
reticulate::use_virtualenv(path_venv)
```


If you prefer to store your virtual environment in the project folder, `reticulate` will automatically discover it, and there is no need to call `use_virtualenv()`.


## Usage

### Initialize a BirdNET model
To begin using the BirdNET model, it must first be initialized. During this step, the required model is downloaded if necessary, loaded into memory, and prepared for making predictions.

Several model variations are available, including the TensorFlow Lite model, which is smaller and more lightweight, and the Protobuf model, which is larger but capable of running on GPU hardware for faster performance.

You can also load a custom model if one is available. For information on training custom models, please refer to the BirdNET-Analyzer repository.

```{r init_model}
# The models are defined using the birdnet_model_* family of functions.
# See ?birdnet_model_load for more details.

# Initialize the TensorFlow Lite model
birdnet_model_tflite("v2.4")

# Initialize the Protobuf model
birdnet_model_protobuf("v2.4")


```

To load a custom model, provide the path to the folder containing the model files and the classifier name.
Custom classifiers are still based on a specific version of the BirdNET model, so you need to specify the version as well.

```{r init_custom_model}
classifier_folder <- "/path/to/custom/model"
classifier_name <- "Custom_Classifier"

birdnet_model_custom("v2.4", classifier_folder = classifier_folder, classifier_name = classifier_name)

```



### Identify species in an audio file
With BirdNET, you can identify bird species present in an audio file. The function returns predictions for each 3-second snippet of the audio that exceeds the specified min_confidence threshold. Each row in the resulting data frame represents a single prediction for a specific 3-second interval.

```{r species_in_audio}
library(birdnetR)

# Initialize the TFLite BirdNET Model
model <- birdnet_model_tflite("v2.4")

# Path to an example audio file (replace with your own file path)
audio_path <- system.file("extdata", "soundscape.wav", package = "birdnetR")

# Predict species in the audio file
predictions <- predict_species_from_audio_file(model, audio_path, min_confidence = 0.3, keep_empty = FALSE)

# Example output:
#   start end      scientific_name            common_name confidence
#       0   3 Poecile atricapillus Black-capped Chickadee  0.8140557
#       3   6 Poecile atricapillus Black-capped Chickadee  0.3082857
#       9  12 Haemorhous mexicanus            House Finch  0.6393781
#      18  21  Cyanocitta cristata               Blue Jay  0.4352708
#      18  21  Clamator coromandus Chestnut-winged Cuckoo  0.3225890
#      21  24  Cyanocitta cristata               Blue Jay  0.3290859
# ...

```

If there are multiple predictions above the confidence threshold for the same time interval, you will see multiple rows for that interval. To keep only the most probable prediction per interval, you can use the package's convenience function.

```{r top_predictions}
# Get the top prediction for each interval
get_top_prediction(predictions)

# Example output:
#   start end      scientific_name            common_name confidence
#       0   3 Poecile atricapillus Black-capped Chickadee  0.8140557
#       3   6 Poecile atricapillus Black-capped Chickadee  0.3082857
#       9  12 Haemorhous mexicanus            House Finch  0.6393781
#      18  21  Cyanocitta cristata               Blue Jay  0.4352708
#      21  24  Cyanocitta cristata               Blue Jay  0.3290859

# Note: Fewer rows appear for the interval 18-21 as only the top prediction is retained.

```


### Using a custom species list
In many cases, you may not need to identify all 6,000+ species available in the model. To focus on species relevant to your project, you can use a custom species list containing only the necessary class labels. Providing a custom species list will limit the output to that set of species.

Class labels follow a specific format, consisting of the scientific name and the common name, separated by an underscore, like this:
```{r class_label_example}
"Accipiter cooperii_Cooper's Hawk"
"Agelaius phoeniceus_Red-winged Blackbird"
```


To create a custom species list, ensure each class label is placed on a separate line in a .txt file. You can refer to the example included in this package or consult the full list of species that BirdNET was trained on. The exact labels are model-specific and can be retrieved using the `labels_path` function. The `read_labels` function can conveniently load the labels from the file.

```{r label_file_paths}
# Retrieve the path to the full list of BirdNET classes.
# Use this as a template for creating your custom species list, but don't modify this file directly.
labels_path(model, language = "en_us")
# /.../birdnet/models/v2.4/TFLite/labels/en_us.txt"

# Path to the example custom species list with a reduced number of species
custom_species_list <- system.file("extdata", "species_list.txt", package = "birdnetR")
read_labels(custom_species_list)

# [1] "Accipiter cooperii_Cooper's Hawk"              "Agelaius phoeniceus_Red-winged Blackbird"
# [3] "Anas platyrhynchos_Mallard"                    "Anas rubripes_American Black Duck"
# [5] "Ardea herodias_Great Blue Heron"               "Baeolophus bicolor_Tufted Titmouse"
# [7] "Branta canadensis_Canada Goose"                "Bucephala albeola_Bufflehead"
# [9] "Bucephala clangula_Common Goldeneye"           "Buteo jamaicensis_Red-tailed Hawk"
# ...

```

To use the custom species list, pass it as an argument to the `predict_species_from_audio_file` function. Since this is just a character vector, you can also pass the vector directly to the function.

```{r use_custom_species_list}
predict_species_from_audio_file(model, audio_path, filter_species = c("Cyanocitta cristata_Blue Jay", "Junco hyemalis_Dark-eyed Junco"), min_confidence = 0.3, keep_empty = FALSE)

# Example output:
#   start end     scientific_name     common_name confidence
#      18  21 Cyanocitta cristata        Blue Jay  0.4352708
#      21  24 Cyanocitta cristata        Blue Jay  0.3290859
#      33  36      Junco hyemalis Dark-eyed Junco  0.4590625
#      36  39      Junco hyemalis Dark-eyed Junco  0.3536855
#      42  45      Junco hyemalis Dark-eyed Junco  0.7375432

```



### Predict species occurence with the meta model
BirdNET includes a Meta Model that predicts the likelihood of bird species occurrence at a specific location and time of year. This function returns a data frame containing class labels and corresponding confidence values, which indicate the probability of species presence. These labels can also be used to create a custom species list for further analysis.

```{r use_meta_model}
# load the meta model
meta_model <- birdnet_model_meta("v2.4")

# predict species occurrence in Ithaca, NY in week 4 of the year
predict_species_at_location_and_time(meta_model, latitude = 42.5, longitude = -76.45, week = 4)

# Example output:
#   label                                       confidence
#   Cyanocitta cristata_Blue Jay                0.92886776
#   Poecile atricapillus_Black-capped Chickadee 0.90332001
#   Sitta carolinensis_White-breasted Nuthatch  0.83232993
#   Cardinalis cardinalis_Northern Cardinal     0.82705086
#   Junco hyemalis_Dark-eyed Junco              0.82440305
#   Zenaida macroura_Mourning Dove              0.80619872
#   Corvus brachyrhynchos_American Crow         0.80580002
#   Dryobates pubescens_Downy Woodpecker        0.79495054
#   Spinus tristis_American Goldfinch           0.72782934
#   Baeolophus bicolor_Tufted Titmouse          0.63683629

```

For more detailed information on how the Meta Model works, refer to the help file: `?predict_species_at_location_and_time`.


### Translating common species names

The birdnetR package allows you to translate common bird species names into several different languages. To check which languages are supported, you can use the following command:
```{r languages}
# supply the version of the BirdNET model you are using
available_languages("v2.4")
```

To output the common names in your preferred language, initialize the model with the language parameter set to your desired language code:
```{r}
birdnet_model_tflite("v2.4", language = "fr")
```

If you want to view the class labels in a specific language, you can retrieve and inspect them using these commands:

```{r labels_language}
labels_path_lang <- labels_path(model, language = "fr")
read_labels(labels_path_lang)

# Example output:
# [1] "Abroscopus albogularis_Bouscarle à moustaches"            "Abroscopus schisticeps_Bouscarle à face noire"            "Abroscopus superciliaris_Bouscarle à sourcils blancs"    
# [4] "Aburria aburri_Pénélope aburri"                           "Acanthagenys rufogularis_Méliphage à bavette"             "Acanthidops bairdi_Bec-en-cheville gris"                 
# [7] "Acanthis cabaret_Sizerin cabaret"                         "Acanthis flammea_Sizerin flammé"                          "Acanthis hornemanni_Sizerin blanchâtre"                  
# [10] "Acanthisitta chloris_Xénique grimpeur"                    "Acanthiza apicalis_Acanthize troglodyte"                  "Acanthiza chrysorrhoa_Acanthize à croupion jaune"        
# [13] "Acanthiza ewingii_Acanthize de Tasmanie"                  "Acanthiza inornata_Acanthize sobre"                       "Acanthiza lineata_Acanthize ridé"                        
```
 
